### 2023-01-04 paper.Rmd
editing

@@

Because the sample variance is basically defined as an average, k-means anticlustering can be employed to equalize the variance across groups, which is the central insight motivating the new k-plus criterion: To obtain an adaption of the $\mathit{SSE}$ that quantifies the degree to which the variance differs between groups, a new variable is computed for each original variable, representing the squared distances to the overall mean of the variable. We obtain a new data matrix $\mathbf{Y}_{N \times P} = \{y_{ij}\}_{N \times P}$ where $y_{ij} = (x_{ij} - \overline{\mathbf{x}}_j)^2$. On the basis of the data matrix $\mathbf{Y}_{N \times P}$ we can define the criterion $\mathit{SSE_{Var}}$ reflecting how strongly the variances of all attributes vary across the $K$ clusters:

@@

Analogously to the standard $\mathit{SSE}$, the centroid values $\overline{y}_j ^{(k)}$ are given as

$$
\overline{y}_j ^{(k)} = \frac{1}{n_k} \sum\limits_{i \in C_k} y_{ij}.
$$

@@

It is unlikely that an exclusive maximization of $\mathit{SSE_{Var}}$ (i.e., equalizing variances) is of interest without also considering the standard $\mathit{SSE}$ (i.e., equalizing means). A combination of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$, however, can be employed to simultaneously optimize both objectives, which is a reasonable idea when striving for overall between-group similarity. The most simple approach for such a bicriterion optimization is the weighted sum approach [@naidu2014; @marler2010]. That is, both criteria are computed independently from each other and are then combined into a single criterion through a weighted sum:

@@

This formula again has the same form as $\mathit{SSE}$ but is computed on different data. In this case, both the original data $x_{ij}$ as well as the squared distances $y_{ij}$ are used as attributes that enter the objective function. That is, computing the k-plus criterion can be done by first computing the values $y_{ij}$, appending these to the original data matrix, and then computing the $\mathit{SSE}$ on the augmented data matrix. Hence, using the unweighted sum approach, k-plus anticlustering reduces to an augmentation of the data matrix, leaving the original k-means criterion unchanged. Note that a problem may arise if the weights $w_1$ and $w_2$ are ignored when optimizing the k-plus objective. Since the computation of the $y_{ij}$ involves squaring, their range of values is expected to be very different from the raw values $x_{ij}$. This imbalance can strongly and unpredictably influence the relative weight of $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ when computing $\mathit{SSE_{kplus}}$. A solution is to standardize all values $x_{ij}$ and $y_{ij}$ to the same scale before computing $\mathit{SSE_{kplus}}$, e.g. via $z$-standardization.

@@

Research on multiobjective optimization indicates that solutions oftentimes fail to simultaneously satisfy several criteria [@ferligoj1992direct]. It is therefore unclear whether optimizing a bicriterion objective such as $\mathit{SSE_{kplus}}$ can yield a partitioning where both constituting criteria (mean and variance) are similar between groups. A motivating example illustrates that in the case of k-plus anticlustering, aiming to simultaneously optimize similarity of means and variances may be feasible. I created 14 data points following a univariate normal distribution ($M = 15$, $SD = 3$) and then generated all `r n_partitions(14, 2)` possible ways to partition the 14 data points into $K = 2$ equal-sized groups. Figure 1 shows the scatterplot of the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ across partitions. The partition that is represented by the squared symbol has the overall best highest $\mathit{SSE}$, but is outperformed by many other partitions with regard to $\mathit{SSE_{Var}}$. Similarly, the partition that maximizes $\mathit{SSE_{Var}}$ (the triangular symbol in Figure 1) is not well-suited to optimize $\mathit{SSE}$ at the same time. Thus, when only considering one of the objectives, the mathematically optimal partitioning fails to establish groups that are similar both with regard to their means and variances. The diamond symbol represents the partition that maximizes $\mathit{SSE_{kplus}}$, which yields `r df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum $\mathit{SSE}$, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global maximum value of $\mathit{SSE_{Var}}$. Thus, the optimal partition according to $\mathit{SSE_{kplus}}$ fulfills the two constituting criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ to an astonishing degree.

```{r, fig.cap = "The relationship between the objectives $SSE$ and $SSE_{Var}$ across all ways to partition $N = 14$ normally distributed random values into $K = 2$ groups. The partition represented by the triangular minimizes the difference in variances; the partition represented by the square minimizes the difference in means; the partition represented by the diamond maximizes $SSE_{kplus}$ and satisfys both criteria to an almost optimal degree.", fig.height = 5, fig.width = 5}

>>>>

# some functions for motivating example

# Variance objective function with changed order of the arguments 
# (can be called by sapply() etc.)
var_objective <- function(clusters, features) {
  anticlust::variance_objective(features, clusters)
}

# Compute features for k-means-extended criterion
squared_from_mean <- function(data) {
  apply(data, 2, function(x) (x - mean(x))^2)
}

@@

Because the sample variance is basically defined as an average---and k-means anticlustering minimizes differences with regard to averages---k-means anticlustering can be employed to equalize the variance across groups. This constitutes the central insight motivating the new k-plus criterion.

To obtain an adaption of the $\mathit{SSE}$ that quantifies the degree to which the variance differs between groups, a new variable is computed for each original variable; the new variable represents the squared distance of each of the original values to the mean of the variable. We obtain a new data matrix $\mathbf{Y}_{N \times P} = \{y_{ij}\}_{N \times P}$ where $y_{ij} = (x_{ij} - \overline{\mathbf{x}}_j)^2$. On the basis of the data matrix $\mathbf{Y}_{N \times P}$ we can define the criterion $\mathit{SSE_{Var}}$ reflecting how strongly the variances of all attributes vary across the $K$ clusters:

@@

@@

It is unlikely that an exclusive maximization of $\mathit{SSE_{Var}}$ (i.e., equalizing variances) is of interest without also considering the standard $\mathit{SSE}$ (i.e., equalizing means). A combination of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$, however, can be employed to simultaneously optimize both criteria, which is a reasonable idea when striving for overall between-group similarity. The most simple approach for such a bicriterion optimization is the weighted sum approach [@naidu2014; @marler2010]. That is, both criteria are computed independently from each other and are then combined into a single criterion through a weighted sum:

@@

This formula again has the same form as $\mathit{SSE}$ but is computed on different data. In this case, both the original data $x_{ij}$ as well as the squared distances $y_{ij}$ are used as attributes that enter the objective function. That is, computing the k-plus criterion can be done by first computing the values $y_{ij}$, appending them to the original data matrix, and then computing $\mathit{SSE}$ on the augmented data. Hence, using the unweighted sum approach, k-plus anticlustering leaves the original k-means criterion unchanged. Note that a problem may arise if the weights $w_1$ and $w_2$ are ignored when optimizing the k-plus objective. Because the computation of the new attribute values $y_{ij}$ involves squaring, their range of values is expected to be very different from the raw values $x_{ij}$. This imbalance can strongly and unpredictably influence the relative weight of $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ when computing $\mathit{SSE_{kplus}}$. To ensure an equal weight of both criteria, all attributes ($x_{ij}$ and $y_{ij}$) can be standardize before computing $\mathit{SSE_{kplus}}$, e.g. via $z$-standardization.

@@

Research on multiobjective optimization indicates that solutions oftentimes fail to satisfactorily address several criteria at the same time [@ferligoj1992direct]. It is therefore unclear whether optimizing a bicriterion objective such as $\mathit{SSE_{kplus}}$ can yield a partitioning where both constituting criteria (mean and variance) are similar between groups. A motivating example illustrates that in the case of k-plus anticlustering, aiming to simultaneously optimize similarity of means and variances may be feasible. I created 14 data points following a univariate normal distribution ($M = 15$, $SD = 3$) and then generated all `r n_partitions(14, 2)` possible ways to partition the 14 data points into $K = 2$ equal-sized groups. Figure 1 shows the scatterplot of the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ across partitions. The partition that is represented by the squared symbol has the overall best highest $\mathit{SSE}$, but is outperformed by many other partitions with regard to $\mathit{SSE_{Var}}$. Similarly, the partition that maximizes $\mathit{SSE_{Var}}$ (represented by the triangular symbol) is not well-suited to optimize $\mathit{SSE}$ at the same time. Thus, when only considering one of the objectives, the mathematically optimal partitioning fails to establish groups that are similar both with regard to their means and variances. The diamond symbol represents the partition that maximizes $\mathit{SSE_{kplus}}$, which yields `r df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum $\mathit{SSE}$, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global maximum value of $\mathit{SSE_{Var}}$. Thus, the optimal partition according to $\mathit{SSE_{kplus}}$ fulfills the two constituting criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ to an astonishing degree.

```{r, fig.cap = "The relationship between the objectives $SSE$ and $SSE_{Var}$ across all ways to partition $N = 14$ normally distributed random values into $K = 2$ groups. The partition represented by the triangular minimizes the difference in variances; the partition represented by the square minimizes the difference in means; the partition represented by the diamond maximizes $SSE_{kplus}$ and satisfies both criteria to an almost optimal degree.", fig.height = 5, fig.width = 5}
